Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2022-05-20 analysis/paper_draft2.Rmd
updated discussion, internal notes

First, such incompatibility may occur because of missing historical data. Historical strike data tends to be incomplete at best, because jury-selection records in past cases are hard to get due to court practices and other barriers [@Grosso2017; @Wright2018]. For example, some jurisdictions do not require recording relevant or detailed enough information about prospective jurors who were or could have been struck (e.g., their race or gender). This forces researchers to turn to other data sources, if any, to find that information. And judges and court staff vary in how often they fully complete or maintain the information they are required to record.  This presents an issue for the approach presented here if the missing strike data is missing not at random.

Second, even if that past strike data is complete (no missingness), incompatibility may occur if attorney bias depends on certain trial-level characteristics, such as defendant race or the severity of the criminal offenses charged, that take on one value in the current trial but that vary across the historical trials. For example, suppose an attorney who is more likely to strike Black prospective jurors because the defendant is Black; who has zero strike bias in cases with white defendants; and whose historical strike data comes from past trials with an equal number of Black and white defendants. If so, then bias detection in the current trial will depend on whether we use all the historical data or only the subset of past trials with Black defendants

The simulations further show that bias detection increases with the number of an attorney's strikes per trial or the number of historical trials. This is in part because of our conservative initial prior for the bias parameter ($b \sim N(0,2)$) to accord with the legal burden of proving bias on those who challenge an attorney for strike bias. This accords with @Kadane2021, who recommends a neutral and symmetrical prior for estimating bias in _Batson_ challenges [FIX].

<!-- what about Kadane's recommendation to use multiple priors for sensitivity? -->

Thus, given little historical strike data for an attorney and a low number of strikes per trial, the initial prior will dominate, and,  as a result, the model we use will only detect severe bias.

@@

generated a model that can identify bias and incorporates historical trial information - tested this with simulation
made an app to make the use of the model more accessible - hope is that lawyers will use it in practice

TODO: Add piece about small sample effect and fact that intial prior will dominate in small sample setting. <!-- WHAT to ADD to LAST paragraph in DISCUSSION section? -->

future work could focus on initial prior choice, and measuring compatibility between cur and hist to develope, data driven choice for alpha.

>>>>

First, such incompatibility may occur because of missing historical data. Historical strike data tends to be incomplete at best, because jury-selection records in past cases are hard to get due to court practices and other barriers [@Grosso2017; @Wright2018]. For example, some jurisdictions do not require recording relevant or detailed enough information about prospective jurors who were or could have been struck (e.g., their race or gender). And judges and court staff vary in how often they fully complete or maintain the information they are required to record.  This presents an issue for any approach to detecting strike bias, including the one presented here, if the missing data is missing not at random.

Second, even with complete past strike data (no missingness), incompatibility may occur if attorney bias depends on certain trial-level characteristics (e.g., defendant race, charge severity), that take on one value in the current trial but that vary across the historical trials. For example, suppose an attorney who is more likely to strike Black prospective jurors because the defendant is Black; who has zero strike bias in cases with white defendants; and whose historical strike data comes from past trials with an equal number of Black and white defendants. If so, then bias detection in the current trial with a Black defendant may depend in part on whether we use all the historical data or only the subset of past trials with Black defendants.

These reasons, however, only qualify how lawyers can use the approach here to detect strike bias for _Batson_ challenges. For any attorney, the posterior distribution of the bias parameter can interact with _Batson_ procedure in three ways. First, it can aid the challenger in showing that it has met the initial burden of producing just enough evidence of strike bias to justify shifting the burden onto the striking attorney to provide permissible grounds for the challenged strikes. Second, once the striking attorney proffers those reasons, the trial judge can consider the attorney's estimated bias parameter as relevant to how much to believe those reasons. Third, the trial judge can rely on the attorney's estimated bias parameter to decide the ultimate issue: whether the alleged illegal strike bias is more likely than not to be true.

The simulations further show that bias detection increases with the number of an attorney's strikes per trial or the number of historical trials. This is in part because we chose a conservative initial prior for the bias parameter ($b \sim N(0,2)$) to account for _Batson_ challenger's burden of proof. This initial prior accords with @Kadane2021 [p. 51], who suggests a prior for estimating _Batson_ strike bias that is at least unimodal, symmetric, and not dependent on the data. Given little historical strike data for an attorney and a low number of strikes per trial, this initial prior will dominate, and, as a result, the model we use will only detect severe bias.

@@

In this paper, we discussed a Bayesian approach to estimate attorney strike bias that incorporate historical data on an attorney's use of peremptory strikes in past cases and how to use the power prior to adjust the weight of such historical information. Our simulations showed that how well our approach detects attorney bias depending on the number of past trials in which that attorney used strikes and the number of strikes per trial. Finally, we discussed a prototype software application with which attorneys could use our approach in real time during jury selection to detect attorney strike bias.

<!-- future work could focus on initial prior choice, and measuring compatibility between cur and hist to develope, data driven choice for alpha-->

<!-- what about Kadane's idea of using multiple priors for sensitivity? -->
